**Title: Practical Machine Learning Course Project**

**Author: Anshul Goyal**

**email: anshulgoyal2007@gmail.com**

**Data Credit: Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.**

**http://groupware.les.inf.puc-rio.br/har**

# Introduction
The goal of the project is to predict how well people do exercise on the basis of data collected by wearable devices such as Jawbone Up, Nike FuelBand, and Fitbit. 6 participants were asked to perform barbell lifts correctly and incorrectly in 5 different ways. This information "classe" is the target of this project. Data from accelerometers on the belt, forearm, arm, and dumbell is used as explanatory data.

# Preparing the workbench
Loading all the packages and functions required for the project. 
Set the working directory and loading the data

- dev: Development data
- pred: Prediction data
``` {r, warning = FALSE, results='markup', message=FALSE}
rm(list=ls())
# Loading Packages
library(ggplot2)
library(caret)
library(randomForest)
library(gbm)
library(mlearning)
library(corrplot)

# Functions
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
```

###Load Data
``` {r, warning = FALSE, results='markup', message=FALSE}
# Working Directory
setwd("C:/Users/anshul/Desktop/Education/C_p_m_l/assignment")

#Data Load
dev <- read.csv("pml-training.csv", na.strings = c("NA", ""))
pred <- read.csv("pml-testing.csv", na.strings = c("NA", ""))

```

# Creating train and test data
``` {r, warning = FALSE, results='markup', message=FALSE}
# creating testing(70%) & training(30%)
set.seed(1234)
inTrain <- createDataPartition(y=dev$classe, p=0.7, list=F)
training <- dev[inTrain,]
testing <- dev[-inTrain,]
```

# Cleaning and variable selection
### Removing index and timestamps variables
``` {r, warning = FALSE, results='hide', message=FALSE}
### Removing index and timestamps variables
training <- training[,-(1:5)]
testing <- testing[,-(1:5)]
pred <- pred[,-(1:5)]
```

### Removing variables with excessive "NA", removing variables with over 90% missing
``` {r, warning = FALSE, results='hide', message=FALSE}
keep.vars <- (apply(is.na(training), 2, mean) <= 0.9)
training <- training[,keep.vars]
testing <- testing[,keep.vars]
pred <- pred[,keep.vars] #last variable is problem_id
names(training)==names(pred) #check for variable names in development data and prediction data
```

### Removing variables with nero zero variance
``` {r}
N <- nearZeroVar(training,saveMetrics=T)
N[N$nzv==1,]
```

Removing following variable from predictors
- new_window

``` {r}
training <- training[,-1]
testing <- testing[,-1]
pred <- pred[,-1] #last variable is problem_id
```

### Matching the class of variables in training and prediction dataset
``` {r}
type.summ <- matrix(0,dim(training)[2],5)
colnames(type.summ) <- c("num","variable","same_class", "same_type", "type_training")
for (i in 1:dim(training)[2]){
  type.summ[i,1] <- i
  type.summ[i,2] <- names(training)[i]
  type.summ[i,3] <- (class(pred[,i]) == class(training[,i]))
  type.summ[i,4] <- (typeof(pred[,i]) == typeof(training[,i]))
  type.summ[i,5] <- class(training[,i])
}

print(type.summ[type.summ[,3]==F,])
```

Changing the class of variables in prediction dataset to make it consistent with training
- magnet_dumbbell_z
- magnet_forearm_y
- magnet_forearm_z

``` {r}
pred[,40] <- as.numeric(pred[,40])
pred[,52] <- as.numeric(pred[,52])
pred[,53] <- as.numeric(pred[,53])
```

### Correlation of independent variables
``` {r fig.width=12, fig.height=12}
par(mfrow=c(2,2))
M1 <- cor(training[,1:13])
M2 <- cor(training[,14:26])
M3 <- cor(training[,27:39])
M4 <- cor(training[,40:53])
corrplot(M1, method = "circle")
corrplot(M2, method = "circle")
corrplot(M3, method = "circle")
corrplot(M4, method = "circle")
```

# Modeling
Developing random forest with 100 trees and default parameters

### Creating the modeling equation
``` {r, warning = FALSE}
# independent variables: total 53 independent variables
var.ind <- names(training[,-54]) 
#target variable
var.tar <- "classe" 

mdl.eq <- as.formula(paste(var.tar,"~", paste(c(var.ind), collapse = " + "), sep =""))
print(mdl.eq)
```

### Random Forest
``` {r, warning = FALSE, results='hide'}
set.seed(1234)
rf.mdl = randomForest(mdl.eq, data = training, ntree = 100, do.trace = T, importance= TRUE)

# Prediction: classe_pred
training$classe_pred <- predict(rf.mdl,training)
testing$classe_pred <- predict(rf.mdl,testing)
```

### Variable Importance: Top 10 variables
``` {r, warning = FALSE}
var.imp <- rf.mdl$importance
var.imp <- var.imp[order(-var.imp[,6]),]
var.imp.mtr <- row.names(var.imp)[1:10]
data.frame(var.imp.mtr)
```

### Feature plot of top 10 variables
```{r, warning = FALSE, fig.width=9, fig.height=9}
featurePlot(x=training[,c("num_window","magnet_dumbbell_y","roll_belt", "magnet_dumbbell_z", "yaw_belt")],y=training$classe ,plot="pairs")

featurePlot(x=training[,c("roll_forearm","magnet_dumbbell_x","pitch_belt", "pitch_forearm", "roll_dumbbell")],y=training$classe ,plot="pairs")
```

### confusion Matrix
``` {r, warning = FALSE}
print(confusionMatrix(testing$classe_pred,testing$classe))
print(confusionMatrix(testing$classe_pred,testing$classe))
```
**Expected out of sample accuracy is 99.8%**

# Prediction (on out of sample 20 records)
``` {r, warning = FALSE}

pred$classe_pred <- predict(rf.mdl, pred)
pred <- pred[order(pred$problem_id),]
answers = pred$classe_pred
pml_write_files(answers)

```
